% LaTeX source for textbook ``How to think like a computer scientist''
% Copyright (C) 1999  Allen B. Downey



\chapter{Strings and things}
\label{strings}

\section{Different Strings available}

We have seen five types of values---booleans, characters, integers,
floating-point numbers and strings---but only four types of
variables---{\tt bool}, {\tt char}, {\tt int} and {\tt double}.  So
far we have no way to store a string in a variable or perform
operations on strings.

In fact, there are several kinds of variables in C++ that
can store strings.  One is a basic type that is part of the C++
language, sometimes called ``a native C string.''  The syntax
for C strings is a bit ugly, and using them requires using similar
methods as traditional arrays, so for the most part we are going to
avoid them. But, here are the basics:
\section{C-Strings}
\index{C string}
What people call c-strings are just character arrays:
\begin{verbatim}
    char myString[]{ "Hi, I am a string!" };
\end{verbatim}
The array is large enough to hold all the characters in the string and an extra
character at the end. Every native c-string ends with the character '\\0'. This
is why these strings are sometimes called "null-terminated strings." For example,
\begin{verbatim}
    char hellostr[6]{'h', 'e', 'l', 'l', 'o', '\0'};
\end{verbatim}
Although the word "hello" only has 5 letters, the array needs to be at least 6 characters
to hold the extra $'\backslash 0'$.

You can
use all of the methods you learned with built in arrays with the native c-strings.
\section{C++ strings}
The string type we are going to use is called {\tt string}, which is
one of the classes that belong to the C++ Standard Library.
\footnote{You might be wondering what I mean by {\bf class}.It will be a few
more chapters before I can give a complete definition, but for now a
class is a collection of functions that defines the operations we
can perform on some type.  The {\tt string} class contains all
the functions that apply to {\tt string}s.}

Unfortunately, it is not possible to avoid C strings altogether.
In a few places in this chapter I will warn you about some problems
you might run into using {\tt string}s instead of C strings.

\section{{\tt string} variables}

You can create a variable with type {\tt string} in the usual
ways:

\begin{verbatim}
  std::string first;
  first = "Hello, ";
  std::string second = "world.";
\end{verbatim}
%
The first line creates a {\tt string} without giving it a value.
The second line assigns it the string value {\tt "Hello."}
The third line is a combined declaration and assignment, also
called an initialization.

Normally when string values like {\tt "Hello, "} or {\tt "world."}
appear, they are treated as C strings.  In this case, when we assign
them to a {\tt string} variable, they are converted automatically
to {\tt string} values.

We can output strings in the usual way:

\begin{verbatim}
  std::cout << first << second << std::endl;
\end{verbatim}
%

In order to compile this code, you will have to include the
header file for the {\tt string} class, which means you will need to
add the line \#include<string> to your file.  

Before proceeding, you should type in the program above and make
sure you can compile and run it.

\section{Getting strings from user input}
\index{input!keyboard}

{\tt cin} can also be used to input a {\tt string}:

\begin{verbatim}
  std::string name;

  std::cout << "What is your name? ";
  std::cin >> name;
  std::cout << name << std::endl;
\end{verbatim}
%
Unfortunately, this statement only takes the first word of
input, and leaves the rest for the next input statement.
So, if you run this program and type your full name, it
will only output your first name.

In order to get a string with spaces, I use a function in the header {\tt string} called {\tt getline}.

\begin{verbatim}
  std::string name;

  std::cout << "What is your name? ";
  std::getline (std::cin, name);
  std::cout << name << std::endl;
\end{verbatim}
%
The first argument to {\tt getline} is {\tt cin}, which is
where the input is coming from.  The second argument is the
name of the {\tt string} where you want the result to be
stored.

{\tt getline} reads the entire line until the user hits
Return or Enter.  This is useful for inputting strings that
contain spaces.

In fact, {\tt getline} is generally useful for getting input
of any kind.  For example, if you wanted the user to type an
integer, you could input a string and then check to see if
it is a valid integer.  If so, you can convert it to an integer
value.  If not, you can print an error message and ask the user
to try again.

To convert a string to an integer, you can use the {\tt atoi}
function defined in the header file {\tt cstdlib}.  We will
get to that in Section~\ref{parsing}.

\section{Extracting characters from a string}

Strings are called ``strings'' because they are made up of a sequence,
or string, of characters.  The first operation we are going to
perform on a string is to extract one of the characters.  C++
can use square brackets ({\tt [} and {\tt ]}) for this operation:

\begin{verbatim}
  string fruit = "banana";
  char letter = fruit[1];
  cout << letter << endl;
\end{verbatim}
%
You can also use the "at" command that we learned about when we were 
learning about std::arrays in Section ~\ref{stdarray}.
\begin{verbatim}
  std::string fruit = "banana";
  char letter = fruit.at(1);
  std::cout << letter << std::endl;
\end{verbatim}
The expression {\tt fruit[1]} and {\tt fruit.at(1)} both indicate that I want character number 1
from the string named {\tt fruit}.  The result is stored in a {\tt
char} named {\tt letter}.  When I output the value of {\tt letter}, I
get a surprise:

\begin{verbatim}
a
\end{verbatim}
%
{\tt a} is not the first letter of {\tt "banana"}.  Unless you are a
computer scientist.  For reasons I will not get into, computer scientists always
start counting from zero.  The 0th letter (``zeroeth'') of {\tt
"banana"} is {\tt b}.  The 1th letter (``oneth'') is {\tt a} and the
2th (``twoeth'') letter is {\tt n}.

If you want the zereoth letter of a string, you have to put
zero in the square brackets:

\begin{verbatim}
  char letter = fruit[0];
\end{verbatim}
or
\begin{verbatim}
  char letter = fruit.at(0);
\end{verbatim}

\section{Length}
\index{string!length}
\index{length!string}

To find the length of a string (number of characters), we can
use the {\tt length} function.  The syntax for calling this
function is a little different from what we've seen before:

\begin{verbatim}
  int length;
  length = fruit.length();
\end{verbatim}
%
To describe this function call, we would say that we are {\bf
invoking} the length function on the string named {\tt fruit}.  This
vocabulary may seem strange, but we will see many more examples where
we invoke a function on an object.  The syntax for function invocation
is called ``dot notation,'' because the dot (period) separates the
name of the object, {\tt fruit}, from the name of the function, {\tt
length}.

{\tt length} takes no arguments, as indicated by the empty parentheses
{\tt ()}.  The return value is an integer, in this case 6.  Notice
that it is legal to have a variable with the same name as a function.

To find the last letter of a string, you might be tempted to
try something like

\begin{verbatim}
  int length = fruit.length();
  char last = fruit[length];       // WRONG!!
\end{verbatim}
%
That won't work.  The reason is that there is no 6th letter
in {\tt "banana"}.  Since we started counting at 0, the 6
letters are numbered from 0 to 5.  To get the last character,
you have to subtract 1 from {\tt length}.

\begin{verbatim}
  int length = fruit.length();
  char last = fruit[length-1];
\end{verbatim}

Also, if you did this code with the at method, there would have
been an error when you ran the code:
\begin{verbatim}
  int length = fruit.length();
  char last = fruit.at(length);       // WRONG!!
\end{verbatim}

\section{A run-time error}
\index{error!run-time}
\index{run-time error}

Way back in Section~\ref{run-time} I talked about run-time errors,
which are errors that don't appear until a program has started
running.

So far, you probably haven't seen many run-time errors, because we
haven't been doing many things that can cause one.  Well, now we are.
If you use the {\tt at} method and you provide an index that is
negative or greater than {\tt length-1}, you will get a run-time
error and a message something like this:

\begin{verbatim}
index out of range: 6, string: banana
\end{verbatim}
%
Try it in your development environment and see how it looks.


\section{Traversal}
\index{traverse}

A common thing to do with a string is
start at the beginning, select each character in turn, do
something to it, and continue until the end.  This pattern
of processing is called a {\bf traversal}.  A natural
way to encode a traversal is with a {\tt while} statement:

\begin{verbatim}
  int index = 0;
  while (index < fruit.length()) {
    char letter = fruit[index];
    cout << letter << endl;
    index = index + 1;
  }
\end{verbatim}
%
This loop traverses the string and outputs each letter on
a line by itself.  Notice that the condition is
{\tt index < fruit.length()}, which means that when
{\tt index} is equal to the length of the string, the
condition is false and the body of the loop is not executed.
The last character we access is the one with the
index {\tt fruit.length()-1}.

\index{loop variable}
\index{variable!loop}
\index{index}

The name of the loop variable is {\tt index}.  An {\bf
index} is a variable or value used to specify one member of an ordered
set, in this case the set of characters in the string.  The index
indicates (hence the name) which one you want.  The set has to be
ordered so that each letter has an index and each index
refers to a single character.

As an exercise, write a function that takes a {\tt string}
as an argument and that outputs the letters backwards, all on
one line.
\section{Range based for loop}
\index{for loop!range based}
A C++ string is considered a container class, so you can
use a different type of loop - a ranged based for loop. 
We showed an example when we covered std::arrays and std::spans, but we
didn't explain what it was doing. I am a fan of these types of
loops because it reduces the number of things we need to type.

Here is an example of a ranged based loop that does the same
thing as the while loop we saw earlier.
\begin{verbatim}
  for (const auto &letter:fruit){
    std::cout << letter << std::endl;
  }
\end{verbatim}
Let's go into each part of this loop. 
\begin{enumerate}
    \item The "for" lets the code know there will be a for loop
    \item Parentheses surround the information about the loop
    \item The const is letting the code know that the information should not change
    \item auto is an easy way to create a variable. The compile sees what value it
    will be set to, and makes sure it is that type. In this case, std::strings are
    made of characters, so it makes letter be a character.
    \item The \& makes the variable a reference. \footnote{We have not covered references yet. The short version of the explanation is that it is a way to set a value without
    creating more memory. We will cover this in more depth later in Section \ref{reference}}
    \item letter - This is the variable that it is using in the loop. Each time the 
    loop runs, it takes the next character from fruit and places it in letter
    \item : fruit - This is showing where it should be getting the values to 
    put into letter.
\end{enumerate}
This code works because the code knows so much about the "fruit" object. It knows
how many letters are in it, and it knows how to take each character in turn and place it
in letter. No need to keep track of the index. It is doing it for you.

\section{The {\tt find} function}
\index{find}

The {\tt string} class provides several other functions that you can
invoke on strings.  The {\tt find} function is the opposite the
{\tt []} operator.  Instead of taking an index and extracting the
character at that index, {\tt find} takes a character and finds the
index where that character appears.

\begin{verbatim}
  string fruit = "banana";
  int index = fruit.find('a');
\end{verbatim}
%
This example finds the index of the letter {\tt 'a'} in the string.
In this case, the letter appears three times, so it is not obvious
what {\tt find} should do.  According to the documentation, it returns
the index of the {\em first} appearance, so the result is 1.  If the
given letter does not appear in the string, {\tt find} returns "std::string::npos".

In addition, there is a
version of {\tt find} that takes another {\tt string} as
an argument and that finds the index where the substring
appears in the string.  For example,

\begin{verbatim}
  string fruit = "banana";
  int index = fruit.find("nan");
\end{verbatim}
%
This example returns the value 2.

You should remember from Section~\ref{overloading} that there
can be more than one function with the same name, as long as they
take a different number of parameters or different types.  In
this case, C++ knows which version of {\tt find} to invoke
by looking at the type of argument we provide.

\section{Our own version of {\tt find}}

If we are looking for a letter in a {\tt string}, we may
not want to start at the beginning of the string.  One way
to generalize the {\tt find} function is to write a version
that takes an additional parameter---the index where we should
start looking.  Here is an implementation of this function.

\begin{verbatim}
int find (string s, char c, int i)
{
  while (i<s.length()) {
    if (s[i] == c) return i;
    i = i + 1;
  }
  return string::npos;
}
\end{verbatim}
%
Instead of invoking this function on a {\tt string}, like
the first version of {\tt find}, we have to pass the {\tt string}
as the first argument.  The other arguments are the character
we are looking for and the index where we should start.

The above code should give you an idea of how the find function works internally. But you do not have to write your own function. The string object does have a find that works similarly.

\begin{verbatim}
  string fruit = "banana";
  int index = fruit.find('a', 2);
\end{verbatim}
In the above code snippet, it looks for the character 'a' after the letter at the "2" index. This code would return the number 3.

\section{Looping and counting}
\label{loopcount}
\index{traverse!counting}
\index{loop!counting}

The following program counts the
number of times the letter {\tt 'a'} appears in a string:

\begin{verbatim}
  string fruit = "banana";
  int length = fruit.length();
  int count = 0;

  int index = 0;
  while (index < length) {
    if (fruit[index] == 'a') {
      count = count + 1;
    }
    index = index + 1;
  }
  cout << count << endl;
\end{verbatim}
%
This program demonstrates a common idiom, called a {\bf counter}.  The
variable {\tt count} is initialized to zero and then incremented each
time we find an {\tt 'a'}.  (To {\bf increment} is to increase by one;
it is the opposite of {\bf decrement}, and unrelated to {\bf
excrement}, which is a noun.)  When we exit the loop, {\tt count}
contains the result: the total number of a's.
\footnote{Some of you may be saying, "isn't there something in the language that already does this?" Yes, there is. There is an algorithm named count. If you include "algorithm", the following code will count for you --   
std::cout \textless\textless "Number of times 'a' appears : "
         \textless\textless std::count(fruit.begin(), fruit.end(), 'a'); 
The count code was to show an example of a counter. 
}
\index{counter}
\index{increment}
\index{decrement}

As an exercise, encapsulate this code in a function named
{\tt countLetters}, and generalize it so that it accepts the
string and the letter as arguments.

\index{encapsulation}
\index{generalization}

As a second exercise, rewrite this function so that instead
of traversing the string, it uses the version of
{\tt find} we wrote in the previous section.


\section{String concatenation}

Interestingly, the {\tt +} operator can be used on strings;
it performs string {\bf concatenation}.  To concatenate means to
join the two operands end to end.  For example:

\begin{verbatim}
  string fruit = "banana";
  string bakedGood = " nut bread";
  string dessert = fruit + bakedGood;
  cout << dessert << endl;
\end{verbatim}
%
The output of this program is {\tt banana nut bread}.

Unfortunately, the {\tt +} operator does not work on native
C strings, so you cannot write something like

\begin{verbatim}
  string dessert = "banana" + " nut bread";
\end{verbatim}
%
because both operands are C strings.  As long as one of the
operands is  {\tt string}, though, C++ will automatically
convert the other.

It is also possible to concatenate a character onto the
beginning or end of a {\tt string}.  In the following example, we
will use concatenation and character arithmetic to output
an abecedarian series.

``Abecedarian'' refers to a series or list in which the elements
appear in alphabetical order.  For example, in Robert McCloskey's book
{\em Make Way for Ducklings}, the names of the ducklings are Jack,
Kack, Lack, Mack, Nack, Ouack, Pack and Quack.  Here is a loop that
outputs these names in order:

\begin{verbatim}
  string suffix = "ack";

  char letter = 'J';
  while (letter <= 'Q') {
    cout << letter + suffix << endl;
    letter++;
  }
\end{verbatim}
%
The output of this program is:

\begin{verbatim}
Jack
Kack
Lack
Mack
Nack
Oack
Pack
Qack
\end{verbatim}
%
Of course, that's not quite right because I've misspelled ``Ouack''
and ``Quack.''  As an exercise, modify the program to correct
this error.

Again, be careful to use string concatenation only with {\tt string}s
and not with native C strings.  Unfortunately, an expression like
{\tt letter + "ack"} is syntactically legal in C++, although it
produces a very strange result, at least in my development environment.

\section{{\tt string}s are mutable}
\index{class!string}
\index{immutable}
\index{string}

You can change the letters in a {\tt string} one at a time
using the {\tt []} operator on the left side of an assignment.
For example,

\begin{verbatim}
  string greeting = "Hello, world!";
  greeting[0] = 'J';
  cout << greeting << endl;
\end{verbatim}
%
produces the output {\tt Jello, world!}.


\section{{\tt string}s are comparable}
\label{incomparable}
\index{class!string}
\index{comparison!string}
\index{string}

All the comparison operators that work on {\tt int}s and
{\tt double}s also work on {\tt strings}.  For example,
if you want to know if two strings are equal:

\begin{verbatim}
  if (word == "banana") {
    cout << "Yes, we have no bananas!" << endl;
  }
\end{verbatim}
%
The other comparison operations are useful for putting words
in alphabetical order.

\begin{verbatim}
  if (word < "banana") {
    cout << "Your word, " << word << ", comes before banana." << endl;
  } else if (word > "banana") {
    cout << "Your word, " << word << ", comes after banana." << endl;
  } else {
    cout << "Yes, we have no bananas!" << endl;
  }
\end{verbatim}
%
You should be aware, though, that the {\tt string} class does
not handle upper and lower case letters the same way that people
do.  All the upper case letters come before all the lower case
letters.  As a result,

\begin{verbatim}
Your word, Zebra, comes before banana.
\end{verbatim}
%
A common way to address this problem is to convert strings to a
standard format, like all lower-case, before performing the
comparison.  The next section explains how.  I will not address the
more difficult problem, which is making the program realize that
zebras are not fruit.

\section{Character classification}

It is often useful to examine a character and test whether
it is upper or lower case, or whether it is a character or
a digit.  C++ provides a library of functions that perform
this kind of character classification.  In order to use these
functions, you have to include the header file {\tt cctype}.

\begin{verbatim}
  char letter = 'a';
  if (isalpha(letter)) {
    cout << "The character " << letter << " is a letter." << endl;
  }
\end{verbatim}
%
You might expect the return value from {\tt isalpha} to
be a {\tt bool}, but for reasons I don't even want to think
about, it is actually an integer that is
0 if the argument is not a letter, and some non-zero value
if it is.

This oddity is not as inconvenient as it seems, because it is
legal to use this kind of integer in a conditional, as shown
in the example.  The value 0 is treated as {\tt false}, and
all non-zero values are treated as {\tt true}.

Technically, this sort of thing should not be allowed---integers are
not the same thing as boolean values.  Nevertheless, the C++ habit of
converting automatically between types can be useful. \footnote{If you are wondering why the call is like this, this call is originally from the language 'C'. At the time the function was created, the language did not have a bool type variable. The zero/not zero was the only way that they could get a bool like experience. Luckily, the language writers decided to add booleans. But, old functions like this still exist, showing what was available years ago.}

Other character classification functions include {\tt isdigit}, which
identifies the digits 0 through 9, and {\tt isspace}, which identifies
all kinds of ``white'' space, including spaces, tabs, newlines, and a
few others.  There are also {\tt isupper} and {\tt islower}, which
distinguish upper and lower case letters.

Finally, there are two functions that convert letters from one
case to the other, called {\tt toupper} and {\tt tolower}.  Both take
a single character as a parameter and return a (possibly
converted) character.

\begin{verbatim}
  char letter = 'a';
  letter = toupper (letter);
  cout << letter << endl;
\end{verbatim}
%
The output of this code is {\tt A}.

As an exercise, use the character classification and conversion
library to write functions named {\tt stringToUpper} and
{\tt stringToLower} that take a single {\tt string} as
a parameter, and that modify the string by converting all the
letters to upper or lower case.  The return type should be
{\tt void}.

\section{Other {\tt string} functions}

This chapter does not cover all the {\tt string} functions.
Two additional ones, {\tt c\_str} and {\tt substr}, are covered
in Section~\ref{finput} and Section~\ref{parsing}.

\section{string\_view}
In C++17, string\_view was added to the language. String
views are another way that we can use strings. This version
of strings is good for variables that we are not planning on
changing. It is a "view" of a string in memory that is
somewhere else. This type needs the code to "\#include \textless  string\_view \textgreater"

If you are sending a string (or c-string) into a function
and you do not want it to change in the function, using std::string\_view as the parameter will be the preferred solution over the const std::string\& that 
was required before C++17.

\begin{verbatim}
void printme(std::string_view yum)
{
  std::cout << yum << std::endl;
}
\end{verbatim}

String views are just a view of a different variable's memory. If the
other variable was changed, the view variable would be changed as well.

\begin{verbatim}
    std::string hey{"hey "};
    std::string_view nothey{hey};
    std::cout << nothey;
    hey[1]='i';
    std::cout << nothey;
    
This has the output:
hey hiy
\end{verbatim}

\section{Glossary}

\begin{description}

\item[object:] A collection of related data that comes with a set of
functions that operate on it.  The objects we have used so far are the
{\tt cout} object provided by the system, std::arrays, spans and {\tt string}s.

\item[index:]  A variable or value used to select one of the
members of an ordered set, like a character from a string.

\item[traverse:]  To iterate through all the elements of a set
performing a similar operation on each.

\item[counter:]  A variable used to count something, usually
initialized to zero and then incremented.

\item[increment:]  Increase the value of a variable by one.
The increment operator in C++ is {\tt ++}.  In fact, that's
why C++ is called C++, because it is meant to be one better
than C.

\item[decrement:]  Decrease the value of a variable by one.
The decrement operator in C++ is {\tt --}.

\item[concatenate:] To join two operands end-to-end.

\index{object}
\index{index}
\index{traverse}
\index{counter}
\index{increment}
\index{decrement}
\index{concatenate}

\end{description}
